High frequency Intraday Forex rates forecasting using an online cointegration approach

### 2015-06-23 manuscript/manuscript.tex
saving my work

Vector error correction model (VECM) introduces the long-run relationship among a set of cointegrated variables
as an error correction term. VECM is a special case of the vector autoregressive
model (VAR) model. VAR model expresses future values as a linear combination of
variables past values.  However, VAR model cannot be used with non-stationary
variables, which is a common feature of financial assets. VECM is a linear model
but in terms of variable differences. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium. In finance,
many economic time series (foreign exchange market rates for example) turns out to be stationary, when they are differentiated once ($I(1)$) and
cointegration restrictions applied on the model often improves forecasting~\cite{duy1998}. Therefore,
VECM has been widely adopted.

VECM parameters are obtained using ordinary least squares
(OLS) method \cite{golub1980}. OLS has two main problems: is sensitive to errors on input data
and involves many calculations. The former problem is commonly solved using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
parameter that leads to an unbiased estimation with better generalisation
capability. The second problem of computational complexity depends on the number
of past values and observations considered.  Recently, online learning
algorithms have been proposed to solve problems with large data sets because of
their simplicity and their ability to update the model when new data is
available. We use an online version of the linear ridge regression and also a modification introduced by 
Vovk \cite{vovk2001} called the Aggregating algorithm regression (AAR), which is also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}.

Our proposal is an online formulation of the VECM called Online VECM (OVECM)
based on consideration of only a sliding window of the historical data.  OVECM parameters are obtained
via OLS, RR or AAR. A competitive analysis is done comparing its cumulative losses against the optimal offline algorithm.

In finance, little academic research has been done using high-frequency Forex
rates \cite{Bekiros2015}. Our proposal is tested using four
three mayor currency pairs (MCP) from the foreign exchange market:  Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
USD (GBPUSD) and USD to Swiss Franc (CHF) (USDCHF). We used samples with a frequency of 10 seconds.

@@

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation.

@@

The solution $\widehat{\mathbf{A}}$ to
equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
\begin{equation*}
\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
\end{equation*}
\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
\begin{equation*}
\label{eq:MP}
\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
\end{equation*}
\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
which, when $\mathbf{A}$ is full rank, can be written as follows: 
\begin{equation}
\label{eq:pseudoinverse}
\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
\end{equation}
%when $\mathbf{A}$ is not full rank, i.e
%$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
%always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
%More generally, the pseudo-inverse is best computed using the compact
%singular value decomposition (SVD) of $\mathbf{A}$ which is:
%\begin{equation*}
%    \label{eq:compactsvd}
%    \mathbf{A}=
%    \mathbf{U_1}
%    \boldsymbol \Sigma_1
%    \mathbf{V}_1^{\top} \, .
%\end{equation*}
%Pseudo-inverse can then be written as follows:
%\begin{equation*}
%\label{eq:pseudoinversesvd}
%\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
%\end{equation*}

@@

\begin{equation}
\label{eq:RRproblem} 
J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
 \| \mathbf{\mathbf{X}}\| ^2
\end{equation}

\noindent which optimal solution $\mathbf{\mathbf{X}}_*$ is well known:

\begin{equation}
\label{eq:optsolRR}
\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
\end{equation}

@@

In batch learning, the training phase is commonly a computationally expensive
process. Therefore, when new data arrives it can't be included easily to the
model. Moreover, it could happen that we won't have enough time to process the
new data before more data arrives. In online learning algorithms there is no
training phase, but the model is updated and evaluated at every time step. This
model updating is computationally less expensive than a training phase.
Online algorithms allow incremental learning by processing one instance at a
time. This is done updating the current model instead of building the model from
scratch.

>>>>

The vector error correction model (VECM) introduces the long-run relationship among a set of cointegrated variables
as an error correction term. VECM is a special case of the vector autoregressive
model (VAR) model. VAR model expresses future values as a linear combination of
variables past values.  However, VAR model cannot be used with non-stationary
variables, which is a common feature of financial assets. VECM is a linear model
but in terms of variable differences. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium. In finance,
many economic time series turns out to be stationary, when they are differentiated once (see section \ref{sec:coint} for more details). Additionally, cointegration restrictions applied on the model often improves forecasting~\cite{duy1998}. This is the reason why VECM has been widely adopted.

VECM parameters are obtained using ordinary least squares
(OLS) method \cite{golub1980}. OLS has two main problems: it is sensitive to errors on input data
and also involves many calculations. The former problem is commonly solved using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
parameter that leads to an unbiased estimation with better generalisation
capability. The second problem of computational complexity depends on the number
of past values and observations considered.  Recently, online learning
algorithms have been proposed to solve problems with large data sets because of
their simplicity and their ability to update the model when new data is
available. We use an online version of the linear ridge regression and also a modification introduced by 
Vovk \cite{vovk2001} called the Aggregating algorithm regression (AAR), which is also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}.

Our proposal is an online formulation of the VECM called Online VECM (OVECM)
based on the consideration of only a sliding window of the historical data at every step.  OVECM parameters are obtained via OLS, RR or AAR. In order to measure performance, a competitive analysis is done which compares the proposal cumulative loss against the optimal offline algorithm.

In finance, little academic research has been done using high-frequency Forex
rates \cite{Bekiros2015}. Our proposal is tested using
three mayor currency pairs (MCP) from the foreign exchange market:  Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
USD (GBPUSD) and USD to Swiss Franc (CHF) (USDCHF). We used samples with a frequency of 10 seconds.

@@

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{Y} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{Y}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{Y}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation.

@@

The solution $\widehat{\mathbf{A}}$ to
equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
\begin{equation*}
\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
\|\mathbf{A}\mathbf{X}-\mathbf{Y}\|_2^2
\end{equation*}
\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
\begin{equation*}
\label{eq:MP}
\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{Y}
\end{equation*}
\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
which, when $\mathbf{A}$ is full rank, can be written as follows: 
\begin{equation}
\label{eq:pseudoinverse}
\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
\end{equation}
%when $\mathbf{A}$ is not full rank, i.e
%$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
%always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
%More generally, the pseudo-inverse is best computed using the compact
%singular value decomposition (SVD) of $\mathbf{A}$ which is:
%\begin{equation*}
%    \label{eq:compactsvd}
%    \mathbf{A}=
%    \mathbf{U_1}
%    \boldsymbol \Sigma_1
%    \mathbf{V}_1^{\top} \, .
%\end{equation*}
%Pseudo-inverse can then be written as follows:
%\begin{equation*}
%\label{eq:pseudoinversesvd}
%\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
%\end{equation*}

@@

\begin{equation}
\label{eq:RRproblem} 
\widehat{\mathbf{X}} (\lambda) = \underset{\mathbf{X}}{\text{Arg\;min}}  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
 \| \mathbf{\mathbf{X}}\| ^2
\end{equation}

\noindent which optimal solution is well known:

\begin{equation}
\label{eq:optsolRR}
\widehat{\mathbf{X}} (\lambda)=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \, ,
\end{equation}

@@

In batch learning, the training phase is commonly a computationally expensive
process. Therefore, when new data arrives it can't be included easily to the
model. Moreover, it could happen that there is  won't have enough time to process the
new data before more data arrives. In online learning algorithms there is no
training phase, but the model is updated and evaluated at every time step. This
model updating is computationally less expensive than a training phase.
Online algorithms allow incremental learning by processing one instance at a
time. This is done updating the current model instead of building the model from
scratch.
